\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-02-25 honors_thesis/journal_paper/paper.tex
A lot of work on the journal paper

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel adaptation of the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $k$ neighbors; we use 2 neighbors for all of our
experiments.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
all possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

@@

Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In that topology, 2 neighbors are chosen at random from the
swarm for each particle at each iteration.  Because a particle is also always
its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

@@

@@

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation---there are
potentially more than $2(n+1)-1$ possible next positions, increasing the amount
of work that must be done to perform the second iteration at the same time as
the first.

This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
available information about $\noeval{\nset}_{t+2}$ before producing
$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its 
neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
still only $2(n+1)-1$ possible next positions, and the problem is avoided.

\subsection{Comparison with Standard PSO}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:basic-griewank1}
\end{figure}

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.  Our speculative algorithm has a swarm of 30 particles and
  does two iterations per round, while standard PSO algorithms use 240
  processors and do one iteration per round.}
  \label{fig:rastrigin}
\end{figure}

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results of the previous section show that it requires too many
extra evaluations to produce competitive results.  On most functions, the
larger swarm size that can be used in regular PSO leads to better performance
than is found by trying to do speculative evaluation.  However, if we keep the
idea of speculative evaluation while relaxing the requirement of exactly
reproducing the behavior of the original PSO algorithm, we see some very
impressive results.

Should I present results for each improvement individually, or all together, or
both?

@@

In performing speculative evaluation as we have described it, $2(n+1)-1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  We can try to make use of those evaluations instead of
throwing them away.

@@

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2(n+1)-1$.  While we cannot with certainty predict the correct branch every
time, we can look at the statistics of the behavior of particles in general and
perhaps come up with a good guess, reducing the number of necessary speculative
evaluations.

@@

\begin{table}
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neigbhor\\
  \end{tabular}
\end{table}

@@

Row 1 in Table~\ref{tab:branches} corresponds to the phenomon commonly called
stagnation in the literature, but only with a single particle.  As an
interesting aside, we found that complete stagnation practically never occurs
in PSO, though frequently large percentages of the swarm are stagnant.
Iterations where the entire swarm is stagnant average less than 2 in 1000
iterations, unless the swarm converges beyond machine precision.  When graphs
of the best function value flatline, what really has happened is a contraction
of the particles' velocity, not stagnation in the technical sense.

@@

With the other topologies, however, there are a few branches with very high
probabilities, so we can have hope that cutting out some of the evaluations
that have low probability will lead to an increase in overall performance.

@@

One possibility is to leave the particle as it is, not accepting any of the
speculative evaluations, because none of the were correct.  This leads to
particles being at different iterations, as some particles guess correctly
while others do not.  Thus we lose exact compatibility with the original PSO,
though this particular relaxation is nothing new; asynchronous adaptations of
PSO do the same thing.  As an aside, it is equivalent in this case to simply
increment the iteration number of particles which fail to correctly predict
their branch.  This keeps the iteration number constant across all particles,
simplifying the work needed to be done when dynamic topologies are involved.

@@

@@

We need not simply produce speculative children for the next iteration.  We can
view all possible speculative evaluations for a particle as an infinite tree
with branching factor $2(n+1)-1$.  As we have already seen that doing one full
level of the tree produces too many extra evaluations to be profitable, it is
incredibly unlikely that doing two full iterations would produce decent
results.  But, if the idea of speculating more than one iteration ahead is
combined with wisely pruning the possible evaluations based on branch
statistics, we can use just a few extra evaluations to go two or three
iterations ahead on the most likely branches.

@@

@@

If we use the concept of tokens commonly used in multirobot stuff (see, for
example, \dots), we can specify a number of tokens available for particles to
use, where each token corresponds to a speculative evaluation.  Whichever
method is used to which evaluations to do, they are prioritized so that
particles with only one or two tokens perform the most valuable speculative
evaluations.  Particles that perform well keep their tokens, particles that
perform poorly send one token (if they have one) to another random particle in
the swarm.

@@

There are thus two distinct uses for tokens.  First, they could be used to try
to evenly spread extra evaluations due to an incapatible number of particles
and processors.  In cursory experimentation, we found that when only going one
iteration ahead keeping the number of evaluations per particle as even as
possible worked much better than trying to prioritize particles and let some
guess correctly more often because they have more tokens. (horrible explanation\dots)

The other, more compelling reason to use tokens is when speculating many
iterations ahead.  Particles that are going in good directions, as evidenced by
relatively frequent updates to the particle's personal best, keep their tokens
and are able to speculate up to three or four iterations.  Particles that
aren't doing as well lose their tokens and probably contribute to the
exploration of the algorithm, for which speculative evaluation doesn't help
much.

As in the previous section, we can only offer a very preliminary exploration of
the possibilities of using tokens to improve the allocation of speculative
evaluations to particles.

\section{Results}
\label{sec:results}

@@

@@

We have speculated (pardon the pun\dots) that our ``pick the best child''
method simply draws a number of samples from the next iteration of the sampling
distribution of PSO and keeps one of them.  It would then be expected that the
sampling distribution of that method is very similar to the distribution of the
original PSO.  This remains to be proven, or even demonstrated empirically.

We gave a passing glance to some very interesting statistical characteristics
of PSO.  Our motivation was to use the statistics to improve our speculative
algorithm, but it seems there are some nuggets in the numbers that could be
used to improve our understanding of the behavior of PSO in general, and what
role a topology plays.  A potentially very fruitful avenue of future work would
be to further analyze the branch statistics in PSO.  There are some particles
that seem to drive the swarm and others that are stagnant almost the entire
time.  Such information could be used in a TRIBES-like fashion to determine
which particles should be removed from the swarm, perhaps starting them close
to well-performing particles.

@@

>>>>

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel adaptation of the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $n$ neighbors; we use 2 neighbors for all of our
experiments.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Thus we have $2(n+1)-1$, or $2n+1$, evaluations.
Table~\ref{tab:evals} lists all possible combinations of updates that can be
seen by a particle with two neighbors other than itself, which is the case in
all of the topologies we use.  If all of these speculative positions are
evaluated in parallel along with the original, non-speculative position of the
particle at time $t-1$, then one of the speculative positions has to be the
correct one, and two iterations have been evaluated at once.

@@

Because $2n+1$ speculative evaluations must be performed for each particle, the
choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In that topology, 2 neighbors are chosen at random from the
swarm for each particle at each iteration.  Because a particle is also always
its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

@@

\label{sec:distributed}

@@

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation---there are
potentially more than $2n+1$ possible next positions, increasing the amount of
work that must be done to perform the second iteration at the same time as the
first.

This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
available information about $\noeval{\nset}_{t+2}$ before producing
$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its
neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
still only $2n+1$ possible next positions, and the problem is avoided.

\subsection{Comparison with Standard PSO}
\label{sec:results}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.}
  \label{fig:basic-griewank1}
\end{figure}

@@

Because our speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum twice as fast
as the original PSO without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank3.eps}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:basic-griewank3}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.}
  \label{fig:rastrigin}
\end{figure}

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results of the previous section show that in many cases it
requires too many extra evaluations to produce competitive results.  On most
functions, the larger swarm size that can be used in regular PSO leads to
better performance than is found by trying to do speculative evaluation.
However, if we keep the idea of speculative evaluation while relaxing the
requirement of exactly reproducing the behavior of the original PSO algorithm,
we see some very impressive results.

@@

In performing speculative evaluation as we have described it, $2n+1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  We can try to make use of those evaluations instead of
throwing them away.

@@

We show in \fig{sphere-pickbest} the results of using this modification on the
function Sphere.  For ease of labeling, we call the modified algorithm SEPSO
Pick Best.  As can be seen, where SEPSO did not manage to perform as well as
PSO Complete, SEPSO Pick Best handily outperforms it with the same number of
processors.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere2.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing SEPSO Pick Best with previous results.}
  \label{fig:sphere-pickbest}
\end{figure}

Sphere is a function for which very little exploration needs to be done, so
always keeping the best speculative position found makes intuitive sense.  A
less intuitive result occurs when using SEPSO Pick Best with the function
Griewank.  Griewank is a highly deceptive function prone to premature
convergence.  Our intuition on the behavior of picking the best speculative
child is that it would tend to prematurely converge more often than the
original PSO.  However, as seen in \fig{griewank-pickbest}, SEPSO Pick Best
increases the chance of finding the optimum while also reducing the number of
iterations required to converge.  Perhaps SEPSO Pick Best encourages local
exploitation, while the sparse communication of the Ring topology still allows
for global exploration.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing SEPSO Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  While we cannot with certainty predict the correct branch every time,
we can look at the statistics of the behavior of particles in general and
perhaps come up with a good guess, reducing the number of necessary speculative
evaluations.

@@

\begin{table}
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neighbor\\
  \end{tabular}
\end{table}

@@

Row 1 in Table~\ref{tab:branches} corresponds to the phenomenon commonly called
stagnation in the literature, but only with a single particle.  As an
interesting aside, we found that complete stagnation practically never occurs
in PSO, though frequently large percentages of the swarm are stagnant.
Iterations where the entire swarm is stagnant average less than 2 in 1000
iterations, unless the swarm converges beyond machine precision.  When graphs
of the best function value flat-line, what really has happened is a contraction
of the particles' velocity, not stagnation in the technical sense.

@@

With the other topologies, however, there are a few branches with very high
probabilities, so we can have hope that cutting out some of the evaluations
that have low probability will lead to an increase in overall performance.  In
order to implement this kind of pruning, the only change that needs to be made
to \alg{centralized} is in Step 2, where the speculative children are
generated.  Simply generate a subset of the children that would normally be
generated.

@@

One possibility is to leave the particle as it is, not accepting any of the
speculative evaluations, because none of them were correct.  This leads to
particles being at different iterations, as some particles guess correctly
while others do not.  Thus we lose exact compatibility with the original PSO,
though this particular relaxation is nothing new; asynchronous adaptations of
PSO do the same thing.  As an aside, it is equivalent in this case to simply
increment the iteration number of particles which fail to correctly predict
their branch.  This keeps the iteration number constant across all particles,
simplifying the work needed to be done in determining neighbors when dynamic
topologies are involved.

@@

There are a large number of ways to prune speculative evaluations.  We
experimented with several, but present results only for one possible pruning.
The pruning we present uses only the two most probable branches in the Random
topology, as seen in~\ref{tab:stats}, the two branches where the $\nbest$ value
was not updated.  Those branches are also convenient in that no messages are
needed from neighbors in order to produce the positions of the speculative
particles---in distributed frameworks using several rounds of communication
(see \sec{distributed}), one of the rounds of communication can be dropped
entirely.

Because we are now only doing two speculative evaluations per particle, with
240 processors we can have a swarm of size 80, instead of the 30 particles in
previous experiments.

In \fig{sphere-pruned}, we show the results of this simple pruning on the
function Sphere.  As can be seen, the larger number of particles that pruning
allows results in better results when picking the best speculative child.  
When rejecting all speculative children, which we with tongue in cheek call
Social Promotion, no real improvement is seen.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:sphere-pruned}
\end{figure}

When looking at the function Griewank, presented in \fig{griewank-pruned}, we
see different results.  Our intuition about picking the best child is more
correct in this case; while adding 50 more particles to the swarm, Pick Best
with pruning still gets stuck more often than the original Pick Best.  However,
Social Promotion does well in this case, finding the optimum every time and
faster than the original PSO, without the need to increase the number of
processors.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank5.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:griewank-pruned}
\end{figure}

@@

We need not simply produce speculative children for the next iteration.  We can
view all possible speculative evaluations for a particle as an infinite tree
with branching factor $2n+1$.  As we have already seen that doing one full
level of the tree produces too many extra evaluations to be profitable, it is
incredibly unlikely that doing two full iterations would produce decent
results.  But, if the idea of speculating more than one iteration ahead is
combined with wisely pruning the possible evaluations based on branch
statistics, we can use just a few extra evaluations to go two or more
iterations ahead on the most likely branches.

When speculating more than one iteration ahead, the idea of picking the child
that matches the correct branch, and holding back the particle if no child
matches the branch, does not make very much sense, as correct branches can only
be determined for the first iteration.  Thus when speculating more than one
iteration ahead we always pick the child that has the highest value.

@@

The combination of branches we tried uses seven speculative evaluations per
particle, matching the swarm size of the original speculative algorithm.  The
seven evaluations we used corresponded to several iterations of branches 1 and
2 in Table~\ref{tab:branches}.  Two of the evaluations were just one iteration
ahead, four were two iterations ahead, and one was three iterations ahead.  The
evaluations that were one iteration ahead were branches 1 and 2; those that
went two iterations ahead were formed by taking either branch 1 or 2 and then
branch 1 or 2; and the evaluation going three iterations ahead followed branch
1 on all three iterations.

The results when optimizing the function Sphere, presented in
\fig{sphere-manyiters}, are rather astounding.  When 240 processors are
available for use, performing speculative evaluation several iterations ahead
finds values an average of 46 orders of magnitude better than using the
processors to perform the standard PSO algorithm with 240 particles.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere4.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:sphere-manyiters}
\end{figure}

With the function Griewank, the premature convergence problems inherent in
picking the best child are exacerbated when speculating several iterations
ahead.  When the algorithm converges, it converges quicker than any other 
method we tried.  However, it also got stuck and failed to find the optimum
more than any other method.  In order for this method to be consistently useful
on functions like Griewank, it seems that more processors are required.  The
results are shown in \fig{griewank-manyiters}.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank6.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

If we use the concept of tokens commonly used in multirobot stuff (see, for
example, \dots), we can specify a number of tokens available for particles to
use, where each token corresponds to a speculative evaluation.  Whichever
method is used to decide which evaluations to do, they are prioritized so that
particles with only one or two tokens perform the most valuable speculative
evaluations.  Particles that perform well keep their tokens, and particles that
perform poorly send one token (if they have one) to another random particle in
the swarm.

@@

There are thus two distinct uses for tokens.  First, they could be used to try
to evenly spread extra evaluations due to an incapatible number of particles
and processors.  In cursory experimentation, we found that this was not a very
effective use of tokens.  In virtually every case we tried, it was better to
simply change the number of particles to match the number of processors, and 
spread the speculative evaluations evenly throughout the swarm.

The other, more compelling reason to use tokens is to intelligently and
dynamically allocate speculative evaluations to particles that are performing
well.  This really only makes sense when speculating several iterations ahead;
after one or two speculative evaluations, the added benefit from also
evaluating the low-probability branches is negligible.  But by intelligently
allocating speculative evaluations, particles that are going in good directions
move faster by speculating farther with extra tokens, while particles that are
not doing as well lose their tokens.  Those particles likely contribute to the
exploration of the algorithm, and when they find good locations, they start
keeping their tokens, switching roles to become productive particles
themselves.  This approach is very similar in concept to other dynamic swarms,
like the TRIBES method proposed by Clerc (cite).

As in the previous section, we can only offer a very preliminary exploration of
the possibilities of using tokens to improve the allocation of speculative
evaluations to particles.  We used a distributed parallel PSO framework, and in
this setting determining which particles are performing the best globally is
difficult without extensive communication.  Thus we only present results for
the one method of token allocation described above.

The only function for which we saw a benefit when using this technique was
Griewank.  As seen in \fig{griewank-token}, using tokens increased the chance
of finding the optimum to 100\%, while still using far fewer iterations than
standard PSO.  We assume that the relatively poor performance of the technique
on other functions is due to the lack of better token allocation.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank7.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing token-based allocation with previous results.}
  \label{fig:griewank-token}
\end{figure}

@@

We have described a new technique for using processors in parallel PSO to
improve the performance of the algorithm.  To our knowledge, this is the first
time extra processors have been used to do anything in PSO besides increase the
swarm size.  In an increasingly parallel world, such advancements will prove to
be crucial in the continued effectiveness of PSO.

We have detailed how to perform speculative evaluation in PSO in several
different parallel frameworks.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  We have
shown results that conclusively demonstrate the superiority of our techniques
over naively adding particles to the swarm when extra processors are available.

We have shown several different ways to modify the behavior of PSO when doing
speculative evaluation, and, as expected given the No Free Lunch Theorem,
different modifications show varied performance with different functions.
While our exploration of these techniques has been cursory in places, we
believe the results shown are sufficient to demonstrate their promise.

@@

We have speculated that our ``pick the best child'' method simply draws a
number of samples from the next iteration of the sampling distribution of PSO
and keeps one of them.  It would then be expected that the sampling
distribution of that method is very similar to the distribution of the original
PSO.  This remains to be proven, or even demonstrated empirically.

We gave a passing glance to some very interesting statistical characteristics
of PSO.  Our motivation was to use the statistics to improve our speculative
algorithm, but it seems there are some nuggets in the numbers that could be
used to improve our understanding of the behavior of PSO in general, and what
role a topology plays.  A potentially very fruitful avenue of future work would
be to further analyze the branch statistics in PSO across many functions and
topologies, categorizing both the topologies and the functions and discovering
what it is that makes certain topologies work well with certain functions.
There are also some particles that seem to drive the swarm and others that are
stagnant almost the entire time.  Such information could be used in a
TRIBES-like fashion to determine which particles should be removed from the
swarm, perhaps starting them close to well-performing particles.

@@

We gave only a very little attention to the possibilities of token-based
allocation of speculative evaluations to particles.  We imagine that the
benefits of intelligently and dynamically deciding which particles get extra
evaluations could produce very good results.  Future work could explore the use
of a centralized allocator for extra processors or ideas in the multi-robot
task allocation literature, such as auctions, to better assign processors to
particles for speculative evaluations.  While such methods involve extra
communication, with long-running function evaluations the extra overhead of
intelligently allocating resources would likely be negligible.
